Concurrent Programming with CXL

### 2020-06-23 book/paper.tex
Latest

While spinlocks suffer from starvation, it is a uniform random
process and each process has an equal chance of entering the critical
section.  Thus the probability of starvation is exponentially vanishing.
Unfortunately, such is not the case for the
reader/writer solutions that we presented thus far.
Consider this scenario: there are two readers and one writer.  One reader
is in the critical section while the writer is waiting.  Now the
second reader tries to enter and is able to.  The first reader leaves.
We are now in a similar situation as the initial state with one reader
in the critical section and the writer waiting, but it is not the same
reader.  Unfortunately for the writer, this scenario can repeat itself
indefinitely.  So even if neither reader was in the critical section
all of the time, and the second reader arrived well after the writer,
the writer never had a chance.

In this chapter, we will present a version of a reader/writer lock
implementation that solves this type of starvation.
In particular, the probability that either a reader or writer gets starved
will be exponentially diminishing.
We shall call such a solution \emph{fair}
\index{fairness}
(although it does not quite
match the usual formal nor vernacular concepts of fairness).
\glsadd{fairness}

It is difficult if not impossible to make the reader/writer lock solution of
Figure~\ref{fig:rw2lock} fair.  However, the split binary semaphore solution of
Figure~\ref{fig:RWsplitsema} provides much control over whom to pass baton
to.
Figure~\ref{fig:RWfair} contains a version of Figure~\ref{fig:RWsplitsema}
that is a fair solution to the reader/writer problem.  There are two changes
compared to the original version:

@@

The net effect of this is that if there is contention between readers and
writers, than readers and writers end up alternating entering the critical
section.  While readers can still starve other readers and writers can still
starve other writers, they can not starve one another.  And contention for
a semaphore is resolved probabilistically, each reader and each writer will
eventually have a chance to enter the critical section almost surely.

\section*{Exercises}
\begin{enumerate}
\item Write a fair solution to the one-lane bridge problem of Chapter~\ref{ch:sbs}.
\item TBD.  multiple GPU allocator
\end{enumerate}

@@

In \texttt{release\_rlock}, notice that \texttt{notify \&(wcond)} is invoked when there
are no readers left, \emph{without} checking if there are writers waiting to enter.
With Mesa monitors this is ok, because calling \texttt{notify} is a no-op if no process
is waiting.

\texttt{release\_wlock} executes \texttt{notifyAll \&(wcond)} as well as
\texttt{notify \&(wcond)}.
Again, because we do not keep track of the number of waiting readers or writers, we
have to conservatively assume that all waiting readers can enter, or, alternatively,
up to one waiting writer can enter.  So \texttt{release\_wlock} wakes up all
potential candidates.
There are two things to note here.  First, unlike split binary semaphores or Hoare
monitors, where multiple waiting readers would have to be signaled one at a time in a
baton-passing fashion (see Figure~\ref{fig:RWsplitsema}), with Mesa monitors
all readers are awaked in one fell swoop using \texttt{notifyAll}.
Second, both readers and writers are awakened---this is ok because both execute
\texttt{wait} within a \texttt{while} loop, re-checking the condition that they
are waiting for.  So if both type of processes are waiting, either all the readers
get to enter next or one of the writers gets to enter next.

@@

>>>>

While spinlocks suffer from starvation, it is a uniform random
process and each process has an equal chance of entering the critical
section.  Thus the probability of starvation is exponentially vanishing.
We shall call such a solution \emph{fair}
\index{fairness}
(although it does not quite
match the usual formal nor vernacular concepts of fairness).
\glsadd{fairness}

Unfortunately, such is not the case for the
reader/writer solutions that we presented thus far.
Consider this scenario: there are two readers and one writer.  One reader
is in the critical section while the writer is waiting.  Now the
second reader tries to enter and is able to.  The first reader leaves.
We are now in a similar situation as the initial state with one reader
in the critical section and the writer waiting, but it is not the same
reader.  Unfortunately for the writer, this scenario can repeat itself
indefinitely.  So even if neither reader was in the critical section
all of the time, and the second reader arrived well after the writer,
the writer never had a chance.

SBSs allow much control over which thread runs next and is therefore a
good starting point for developing fair synchronization algorithm
In this chapter, we will present two SBS-based fair reader/writer lock
implementations.
The first, in Figure~\ref{fig:RWfair}, is based on Figure~\ref{fig:RWsplitsema}.
There are two differences:

@@

The net effect of this is that if there is contention between readers and
writers, than readers and writers end up alternating entering the critical
section.  While readers can still starve other readers and writers can still
starve other writers, they cannot starve one another.  And since contention for
a semaphore is resolved probabilistically, each reader and each writer will
eventually have a chance to enter the critical section almost surely.

Figures~\ref{fig:RWqueue} and~\ref{fig:RWqtest} present a different approach
in which processes are allowed into the critical region in a purely First
Come First Served manner.  When a reader arrives, it is allowed in only if
there is no writer in the critical section or waiting to enter.  When a writer
arrives, it is allowed in only if there is no reader or writer in the
critical section (and therefore also not waiting to enter).  To maintain the
arrival order, the solution uses a semaphore per process and a queue.
The \texttt{acquire\_} methods take a pointer to
the semaphore of the process as argument.
Each entry in the queue is a pair consisting of a type of access
(\texttt{.read} or \texttt{.write}) and a pointer to the semaphore of the
processs that is waiting.  Method \texttt{V\_one} considers semaphores to vacate
in the order in which they appear in the queue.

\section*{Exercises}
\begin{enumerate}
\item Write a fair solution to the one-lane bridge problem of Chapter~\ref{ch:sbs}.
If you want to use the queue method, you can change the signature of the
methods to enter the bridge, adding an argument that contains a pointer to a
semaphore.
\end{enumerate}

@@

In \texttt{release\_rlock}, notice that \texttt{notify(\&wcond)} is invoked when there
are no readers left, \emph{without} checking if there are writers waiting to enter.
With Mesa monitors this is ok, because calling \texttt{notify} is a no-op if no process
is waiting.

\texttt{release\_wlock} executes \texttt{notifyAll(\&wcond)} as well as
\texttt{notify(\&wcond)}.
Again, because we do not keep track of the number of waiting readers or writers, we
have to conservatively assume that all waiting readers can enter, or, alternatively,
up to one waiting writer can enter.  So \texttt{release\_wlock} wakes up all
potential candidates.
There are two things to note here.  First, unlike split binary semaphores or Hoare
monitors, where multiple waiting readers would have to be signaled one at a time in a
baton-passing fashion (see Figure~\ref{fig:RWsplitsema}), with Mesa monitors
all readers are awaked in one fell swoop using \texttt{notifyAll}.
Second, both readers and writers are awakened---this is ok because both execute
\texttt{wait} within a \texttt{while} loop, re-checking the condition that they
are waiting for.  So if both type of processes are waiting, either all the readers
get to enter next or one of the writers gets to enter next.

@@

\section*{Exercises}
\begin{enumerate}
\item Write a new version of the GPU allocator of Chapter~\ref{ch:bb}.  In this version,
a process is allowed to allocate a set of GPUs and release a set of GPUs that it
has allocated.  Method \texttt{gpuAllocSet(n)} should block until $n$ GPUs are
available, but it should grant them as soon as they are available.
Method \texttt{gpuReleaseSet(s)} takes a set of GPU identifiers as argument.
A process does not have to return all the GPUs it allocated at once.
\item The specification in the previous question makes the solution unfair.
Explain why this is so.  Then change the specification and the solution so that
it is fair.
\end{enumerate}
